
\newcount\draft\draft=1 % set to 0 for submission or publication
\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb,bbm,amsfonts,syntax}
\usepackage[ligature, inference]{semantic}
\usepackage[T1]{fontenc}
\usepackage{mathpartir}
\usepackage{nccmath}
\usepackage{xxx}

\title{Linguine: Linear Algebra Types: \\ Syntax and Semantics}
\author{Dietrich Geisler \and Irene Yoon \and Adrian Sampson}
\date{}

\newcommand{\mat}{\mathsf{mat}_{n_1{\times}n_2}}
\newcommand{\vv}[1]{\mathsf{vec}_{#1}}
\newcommand{\env}[1]{#1,\sigma}
\newcommand{\defas}{\mathrel{::=}}
\newenvironment{leftalign}%
    {\fleqn[5pt]\csname align*\endcsname}%
    {\csname endalign*\endcsname\endfleqn}
\newcommand{\alt}{\:|\:}

\mathlig{->}{\rightarrow}
\mathlig{|-}{\vdash}
\mathlig{=>}{\Rightarrow}

\begin{document}
\maketitle

\mathligson

\section{Syntax}

Literals in our language can be scalar numbers, vectors, or matrices:
%
\begin{leftalign}
s &\in \mathbb{R} \\
v_n &\defas [s_1,s_2,\dots,s_n] \\
m_{n_1\times n_2} &\defas [[s_{11},s_{12},\dots,s_{1n_2}],\dots,[s_{n_11},s_{n_12},\dots,s_{n_1n_2}]]
\end{leftalign}
%
Expressions can be variables, literals, or operators:
%
\begin{leftalign}
x &\in \text{variables} \\
c &\defas s \alt v_n \alt m_{n_1\times n_2} \\
e &\defas c \alt
    \tau\;x=e \alt
    x=e \alt
    e_1+e_2 \alt
    e_1*e_2 \alt
    e_1\;\mathsf{.*}\;e_2
\end{leftalign}
%
The $.*$ operator represents elementwise multiplication (whereas $*$ is proper linear algebra multiplication).
Our type system consists of user-defined \emph{tags}, $T$, alongside built-in types:
%
\begin{leftalign}
n &\in \mathbb{N} &
\nu &\defas \top_n \alt
\bot_n \alt T \\
T &\in \text{tags} &
\tau &\defas \mathsf{unit} \alt
\mathsf{scalar} \alt
\nu \alt
\nu_1->\nu_2
\end{leftalign}
%
Function types $\nu_1 -> \nu_2$ are not ordinary functions; they are matrices used to map from one vector space to another using matrix--vector multiplication.  As a convenience (as well as to match expected GLSL behavior), the top type $\top_n$ can be written in code as $\mathsf{vec}_n$ and the $\top_{n_1}->\top_{n_2}$ type can be written in code as $\mathsf{mat}n_1\mathsf{x}n_2$.  The $\bot_n$ type is purely meant for use by the typechecker and so is not part of the surface syntax of the language.  

\section{Subtype Ordering}
We define a judgment $\tau_1 \leq \tau_2$ for subtyping on ordinary types.
Subtyping is reflexive and transitive, as usual:
%
\begin{mathpar}
\inferrule
    { }
    {\tau\leq\tau}

\inferrule
    {\tau_1 \leq \tau_2 \and \tau_2 \leq \tau_3}
    {\tau_1 \leq \tau_3}
\end{mathpar}
%
The null type \textsf{unit} acts as a top type for all of $\tau$:
%
\begin{mathpar}
\inferrule
    { }
    {\tau\leq\mathsf{unit}}
\end{mathpar}
%
We also introduce a partial order on tags $\nu$, of the form $\Delta |- \nu_1 \leq_\Delta \nu_2$.  This ordering refers to a context $\Delta$, which is a map from tags $T$ to matrix types $\nu$:
%
\begin{mathpar}
\inferrule
    {\Delta|-\Delta(\nu_1)=\nu_2}
    {\Delta|-\nu_1\leq_\Delta \nu_2}
    \quad \Delta(T)=\nu
\end{mathpar}
%
This relation has the usual reflexivity and transitivity properties:
%
\begin{mathpar}
\inferrule
    { }
    {\Delta|-\nu\leq_\Delta \nu}

\inferrule
    {\Delta|-\nu_1 \leq_\Delta \nu_2 \and \Delta|-\nu_2 \leq_\Delta \nu_3}
    {\Delta|-\nu_1 \leq_\Delta \nu_3}
\end{mathpar}
%
Members of $\nu$, namely vectors and tagged vectors, have a top type representing general vectors of dimension $n$ as $\top_n$.  Additionally, we also introduce a bottom type $\bot_n$ for each vector dimension.  We naturally have the simple rule
%
\begin{mathpar}
\inferrule
    { }
    {\Delta|-\bot_n\leq_\Delta\top_n}
\end{mathpar}
%
The map $\Delta$ should not put anything above the top types $\top_n$ and the bottom types $\bot_n$.  In particular, we require the rules
%
$$\Delta(\nu)\neq\bot_n
\qquad\textnormal{or}\qquad
\Delta(\top_n)\neq\nu$$
%
Additionally, we need a rule to subsume the bottom type into values correctly since $\Delta$ only 'points' towards the top type.  In particular, we require the rule:
\begin{mathpar}
\inferrule
    {\Delta|-\nu\leq_\Delta\top_n}
    {\Delta|-\bot_n\leq_\Delta\nu}
\end{mathpar}
%
This rule, along with the constructions previously described, gives the standard properties of a top and bottom.  In particular, for any $\nu,\Delta$, there exists a $\top_n$ and $\bot_n$ such that
%
\begin{mathpar}
\inferrule
    {}
    {\Delta|-\bot_n\leq_\Delta\nu\quad\textnormal{and}\quad\Delta|-\nu\leq_\Delta\top_n}
\end{mathpar}
%
The function-like transformation matrix types, $\nu_1 -> \nu_2$, have the same standard subtyping relationship as ordinary function types:
%
\begin{mathpar}
\inferrule
	{\Delta|-\nu_1'\leq_\Delta \nu_1\qquad\Delta|-\nu_2\leq_\Delta \nu_2'}
	{\Delta|-\nu_1->\nu_2\leq\nu_1'->\nu_2'}
\end{mathpar}
As with vectors, we can define matrix functions have a $\top_{n_1\times n_2}$ and $\bot_{n_1\times n_2}$ for any pair of dimensions $n_1,n_2$.  These constructions are just maps $\nu_1->\nu_2$, namely:
%
\begin{mathpar}
\inferrule
	{}
	{\top_{n_1\times n_2}=\bot_{n_1}->\top_{n_2}}

\inferrule
	{}
	{\bot_{n_1\times n_2}=\top_{n_1}->\bot_{n_2}}
\end{mathpar}
%
As with vectors, these rules produce the usual property of lattices, namely that every map is 'between' a given pair top and bottom values.  In particular, for any $\nu_1,\nu_2,\Delta$, there exists a $\top_{n_1\times n_2}$ and $\bot_{n_1\times n_2}$ such that:
%
$$\Delta|-\bot_{n_1\times n_2}\leq\nu_1->\nu_2
\quad\textnormal{and}\quad
\Delta|-\nu_1->\nu_2\leq\top_{n_1\times n_2}$$
%
Additionally, for any $\nu_1,\nu_2,\Delta$, there is no $\top_{n_1\times n_2}$ or $\bot_{n_1\times n_2}$ such that
%
$$\Delta|-\top_{n_1\times n_2}\leq\nu_1->\nu_2
\quad\textnormal{or}\quad
\Delta|-\nu_1->\nu_2\leq\bot_{n_1\times n_2}$$
%
\section{Static Semantics}

This typing judgement is a map from an expression to the type of that expression under some variable context $\Gamma$, from variable names to types, and some tag context $\Delta$ defined above.

\subsection{Subsumption}
Any type in a given context can be cast ``up'' at any time.
%
\begin{mathpar}
\inferrule
	{\tau_1\leq\tau_2\qquad\Gamma,\Delta|-e:\tau_1,\Gamma}
	{\Gamma,\Delta|-e:\tau_2,\Gamma}

\inferrule
	{\Delta|-\nu_1\leq_\Delta \nu_2\qquad\Gamma,\Delta|-e:\nu_1,\Gamma}
	{\Gamma,\Delta|-e:\nu_2,\Gamma}
\end{mathpar}
%
\subsection{Constants and Variable Declarations}
Declaring constants produce the types one would expect:
%
\begin{mathpar}
\inferrule
	{ }
	{\Gamma,\Delta|-():\mathsf{unit},\Gamma}

\inferrule
	{ }
	{\Gamma,\Delta|-s:\mathsf{scalar},\Gamma}
\end{mathpar}

Vector and matrix literals take on their respective bottom types of the appropriate dimension (note the swap in dimensions for the bottom type as a result of treating vectors as column vectors).
%
\begin{mathpar}
\inferrule
	{ }
	{\Gamma,\Delta|-v_n:\bot_n,\Gamma}

\inferrule
	{ }
	{\Gamma,\Delta|-m_{n_1\times n_2}:\bot_{n_2\times n_1},\Gamma}
\end{mathpar}

Variables can be declared and assigned.  Assignment is required at declaration time.  Note that the entire variable must be reassigned in the case of vectors and matrices:
%
\begin{mathpar}
\inferrule
	{\Gamma,\Delta|-e:\tau,\Gamma'}
	{\Gamma,\Delta|-\tau\;x:=e:\mathsf{unit}, \Gamma', x \mapsto \tau}

\inferrule
	{\Gamma,\Delta|-\Gamma(x):\tau\qquad\Gamma,\Delta|-e:\tau,\Gamma'}
	{\Gamma,\Delta|-x:=e:\mathsf{unit}, \Gamma'}
\end{mathpar}

\subsection{Binary Operations}

All operators on scalars work as one might expect.

Types are closed under addition and scalar multiplication
%
\begin{mathpar}
\inferrule
	{\Gamma,\Delta|-e_1:\tau,\Gamma'\qquad\Gamma',\Delta|-e_2:\tau,\Gamma''}
	{\Gamma,\Delta|-e_1+e_2:\tau,\Gamma''}
        \quad\tau\neq\mathsf{unit}

\inferrule
	{\Gamma,\Delta|-e_1:\tau,\Gamma'\qquad\Gamma',\Delta|-e_2:\mathsf{scalar},\Gamma''}
	{\Gamma,\Delta|-e_1*e_2:\tau,\Gamma'}
\end{mathpar}

Component multiplication can be defined as a mathematical operation, but makes little formal sense.  Such operations are allowed, but don't interact with spaces and tags and result in a complete lack of information about a resulting matrix.
%
\begin{mathpar}
\inferrule
	{\Gamma,\Delta|-e_1:\top_n,\Gamma'\qquad\Gamma',\Delta|-e_2:\top_n,\Gamma''}
	{\Gamma,\Delta|-e_1\;\mathsf{.*}\;e_2:\top_n,\Gamma''}

\inferrule
	{\Gamma,\Delta|-e_1:\top_{n_1\times n_2},\Gamma'\qquad\Gamma',\Delta|-e_2:\top_{n_1\times n_2},\Gamma''}
	{\Gamma,\Delta|-e_1\;\mathsf{.*}\;e_2:\top_{n_1\times n_2},\Gamma''}
\end{mathpar}

Matrix multiplication is both a way of transforming from one tag to another and for composing two matrix functions together.  Note that these operations are \textit{not} commutative and vectors are treated as column vectors with regards to matching dimensions.
%
\begin{mathpar}
\inferrule
	{\Gamma,\Delta|-e_1:\nu_1->\nu_2,\Gamma'\qquad\Gamma',\Delta|-e_2:\nu_1,\Gamma''}
	{\Gamma,\Delta|-e_1*e_2:\nu_2,\Gamma''}

\inferrule
	{\Gamma,\Delta|-e_1:\nu_2->\nu_3,\Gamma'\qquad\Gamma',\Delta|-e_2:\nu_1->\nu_2,\Gamma''}
	{\Gamma,\Delta|-\;e_1*e_2:\nu_1->\nu_3,\Gamma''}
\end{mathpar}

\section{Dynamic Semantics}

The operational semantics of this language map, in a single step, from a command to a constant and a state $\sigma$.  $\sigma$ itself is, as usual, a map from variable names to constants.

\subsection{Substitution}
Substitutions work as expected on the environment $\sigma$.
%
\begin{mathpar}
\inferrule
	{ }
	{\env{\tau\;x:=c}->(),\sigma[c/x]}

\inferrule
	{\env{e}->\env{e'}}
	{\env{\tau\;x:=e}->\env{\tau\;x:=e'}}
\end{mathpar}

\subsection{Mathematical Operations}
As usual, we can reduce on either side of the mathematical operations $\odot\in\{+,*,\mathsf{.*}\}$
%
\begin{mathpar}
\inferrule
	{\env{e_1}->\env{e_1'}}
	{\env{e_1\odot e_2}->\env{e_1'\odot e_2}}

\inferrule
	{\env{e_2}->\env{e_2'}}
	{\env{c\odot e_2}->\env{c\odot e_2'}}
\end{mathpar}

Addition and scalar multiplication have standard mathematical meaning.  Similarly, vector and matrix multiplication have standard mathematical meaning.  Component-wise multiplication is identical to matrix addition, except using multiplication of numbers in place of addition of numbers.  For readability, formal semantics of each of these operations will be ommitted.
\end{document}
